\chapter{Implementation and code generation}
\label{chapter:process:implementation}

Folderol\footnote{\url{https://github.com/amosr/folderol}} is a Haskell implementation of Machine Fusion \REFTODO{fusion}.
Machine Fusion uses the topology of the entire process network to perform fusion.
When fusing one producer with multiple consumers, the fusion algorithm must coordinate between all the consumers to reach consensus on when to pull the next value.
This coordination between all consumers means the fusion algorithm requires global knowledge of the process network.
In contrast, shortcut fusion systems \REFTODO{shortcut} use rewrite rules to remove intermediate buffers and require only local knowledge, but cannot coordinate between multiple consumers.
In cases where shortcut fusion cannot fuse it fails silently, leaving the programmer unaware of the failure.
This silence is also due to the local nature of rewrite rules: if we wish to know whether all the processes have been fused, we need to know about all the processes.
To fuse the entire process network, as well as to inform the programmer when fusion fails, Folderol uses Template Haskell, a form of metaprogramming.

This chapter looks at the implementation of Folderol, in particular how to generate efficient code for a fused process.
Generating code for a single process is fairly straightforward in itself: the process language is a simple imperative language, and once the process network has been fused into a single process there is no longer any need for threading or inter-process communication.
However, code generation needs to be tailored specifically to take advantage of the optimisations in the target compiler, which in this case is the Glasgow Haskell Compiler (GHC).
Haskell does support imperative constructs like mutable references, but being a functional language, the optimisations in GHC are more geared towards functional programs.
Mutable references are not optimised particularly well.
If we wish to generate efficient code, we must --- perhaps surprisingly --- avoid mutable references, and generate code closer to what the compiler expects.

\input{copy/03-body/04-extraction/s-th.tex}
\input{copy/03-body/04-extraction/s-constructing.tex}
\input{copy/03-body/04-extraction/s-boxing.tex}
\input{copy/03-body/04-extraction/s-endpoints.tex}
\input{copy/03-body/04-extraction/s-codegen.tex}
\input{copy/03-body/04-extraction/s-vector-endpoints.tex}
\input{copy/03-body/04-extraction/s-related.tex}

